---
title: "Portfolio Theory: Chapter 2"
author: "Matt Brigida"
date: "November 27, 2015"
output: html_document
runtime: shiny
---

In this chapter we decide the proportions of our portfolio which we will invest in the risky vs risk free asset.  It may seem odd to simply assume we have two assets in which we'll invest (risky and risk-free) and try and decide the proportions in each.  Why this decision is relevant comes from the next chapter where we construct optimal risky portfolios.

The idea of modern investing is that, because the construction of optimal *risky* portfolios is not dependent on each investors' risk aversion, the construction of risky portfolios can be delegated to professionals.  


Because the proportions in your complete portfolio will be a function of your degree of risk aversion, we'll start with a discussion on terms and our assumptions regarding our behavior towards risk.

## Risk Aversion

We can classify preferences toward risk into three categories: risk averse; risk neutral; risk seeking. 

| Category |  Description          | Type of Person |
|----------|-----------------|-----------------|
| Risk Averse | Rational behavior which only takes on risk for increase expected return. | Investors/Speculators |
| Risk Neutral |  Risk is treated as irrelevant. Assumption often used in academic analyses. | |
| Risk Seeking | Prefers increased risk even if there is no increased expected return. | Gamblers | 

We assume throughout the text that investors are risk averse.  This is an assumption made by the models which make up the basis of portfolio theory---namely Markowitz optimal portfolios and the Capital Asset Pricing Model. 

## Utility Values

So when evaluating a possible investment, an investor prefers the investment the higher its expected return is relative to its risk. Consider the following two investments (say A and B): 

| Investment | Expected Return | Standard Deviation |
|-----|------|------|
| A | 10\% | 20\% |
| B | 10\% | 30\% |

It is easy to see a risk averse investor will choose A.  However what if the investor were confronted with the choice between the following two investments (say C and D):

| Investment | Expected Return | Standard Deviation |
|-----|------|------|
| C | 10\% | 20\% |
| D | 18\% | 30\% |

Now it isn't clear which investment the investor would choose.  Is an additional 8\% expected return worth the increased risk? Maybe and maybe not.  Two risk averse investors could disagree.  We need a systematic way for an investor to rank investments.  For this we use `utility functions'. 

## Utility Functions

We use utility functions to map from the mean/variance space to a number with which we can rank portfolios (higher values are preferred).  We'll use the following utility function:

$U = E(r_X) - \frac{1}{2}A\sigma^2$

$A$ is the degree of risk aversion, and it is particular to each individual.  Thus different investors will have different utility values for the same portfolio.  Note the utility value $U$ is increasing in expected return $E(r)$ and decreasing in risk $\sigma$.

So, say an investor has a degree of risk aversion of 3, and a portfolio has an expected return and standard deviation of 10\% and 17\% respectively.  Then the investor's utility value for the portfolio is: $U = 10\% - \frac{3}{2}(17\%)^2 = 5.67\%$.


```{r calculateUtility, echo=FALSE}
inputPanel(
  sliderInput("expRet", "Expected Return", min = -.2, max = 1, value = .15),
  sliderInput("stDev", "Standard Deviation", min = .01, max = 1, value = .2),
  sliderInput("RA", "Degree of Risk Aversion", min = -.5, max = 6, value = 4)
)
renderPrint({
  U = input$expRet - 0.5 * input$RA * input$stDev^2
  out <- paste(round(U * 100, 2), "%", sep = "")
  cat(out)
})
```


Note, we can find all $(E(r), \sigma)$ pairs such that their utility values are equivalent.  This pairs create *indifference curves* in $(E(r), \sigma)$.  The idea is we don't care where we are on this curve; all points provide us with the same worth.


#### Plot an Indifference Curve
```{r indifferenceCurves, echo=FALSE}
## change to a htmlwidget ----
library(metricsgraphics)
inputPanel(
  sliderInput("utility", "Utility Value", min = .01, max = 1, value = .08),
  sliderInput("RA1", "Degree of Risk Aversion", min = -.5, max = 6, value = 2)
)
## renderPlot({
renderMetricsgraphics({
  ## for each standard deviation, find the expected return which solves for the utility value
  expRet1 <- 0
  index <- seq(.01, 1, by = .01)
 for (i in seq_along(index)) {
  fun <- function(X){ ((X - 0.5 * input$RA1 * index[i]^2) - input$utility)^2 }
  expRet1[i] <- optim(.1, fun, method = "Brent", lower = -1, upper = 5)$par
 }
    tmp <- data.frame(Standard_Deviation = index, Expected_Return = expRet1) %>%
        mjs_plot(x = Standard_Deviation, y = Expected_Return) %>%
        mjs_line() %>%
        mjs_labs(x="Standard Deviation", y="Expected Return")  %>%
        mjs_add_legend(legend=c("Indifference Curve"))
  ## plot(index, expRet1, type = 'l', xlab = "Standard Deviaion", ylab = "Expected Return", lwd = 4, col = 3)
  ## legend("topleft", legend = paste("Utility = ",input$utility, sep = ""), col = 3, lty = 1, lwd = 4)
})
```

### Certainty Equivalence

The utility value is the risk-free rate of return such that we are indifferent between the risky and risk free portfolio.  For example, consider again an investor has a degree of risk aversion of 3, and a portfolio has an expected return and standard deviation of 10\% and 17\% respectively.  The investor's utility value for the portfolio is 5.67\%.  What this means is the investor is indifferent between earning 5.67\% risk free, or investing in a portfolio with a 10\% expected return and 17\% standard deviation. 

So the utility function converts all risky portfolios into thier risk-free counterpart.  This is why we can compare portfolios by their utility value---the utility function adjusts each portfolio for its risk and puts them all on the same scale.  Since the utility value of the risk-free portfolio is simply the risk-free rate, we can also say we'll only consider investing in a risky portfolio if the portfolio's utility value (certainty-equavalent rate) is above the risk-free rate.

# A Portfolio of One Risky and One Risk-Free Asset

Here we assume we already know our optimal risky portfolio, and we are trying to decide how much to invest in this portfolio versus the risk free asset.  Our result is going to be dependent on our particular degree of risk aversion, and therefore each investor will have their own solution.  The construction of the optimal risky portfolio, in the next chapter, will be not be dependent on risk aversion.

The return on an investor's complete portfolio (risky and risk free asset) is:

$r_c = yr_r + (1-y)r_f$

where $r_c$, $r_r$, and $r_f$ are the returns on the complete, risky, and risk free assets respectively.  $y$ is the proportion of your portfolio in the risk-free asset (it can, in general, be negative or greater than 1).  

#### Expected return of the complete portfolio

Taking the expectation and rearranging affords:

$E(r_c) = r_f + y[E(r_r) - r_f]$

So your expected return on your complete portfolio is the risk-free rate plus the product of the risk premium on the risky asset and the proportion you invested in the risky asset.  Note, since investors are assumes risk averse, we know $E(r_r) - r_f > 0$.

#### Standard deviation of the complete portfolio

Noting that the risk-free rate is a constant, the variance if the complete portfolio is $\sigma_c^2 = y^2\sigma_r^2$.  So the standard deviation is $\sigma_c = y\sigma_r$

#### Plotting the complete portfolio

We can rearrange the standard deviation to show $y = \frac{\sigma_c}{\sigma_r}$.  Plugging this in to our equation for the expected return of the complete portfolio yields:

$E(r_c) = r_f + \sigma_c\frac{E(r_r) - r_f}{\sigma_r}$

This allows us to plot our possible $E(r_c)$ and $\sigma_c$ combinations (our investment opportunity set).  We can invest anywhere on the line below by varying $y$, the proportion we invest in the risky asset.  If you allocate less than 100\% of your portfolio in the risky asset, then you invest the remaining portion of your portfolio at the risk-free rate.  However if you have over 100\% of your portfolio in the risky asset, then you borrow at the higher margin rate. The line representing our portfolio opportunity set is known as the Capital Allocation Line (CAL).

```{r opportunitySet, echo=FALSE}
## change to a htmlwidget ----
inputPanel(
  sliderInput("riskfree2", "Risk-Free Rate", min = .01, max = .2, value = .02),
  sliderInput("borrow", "Borrowing (Margin) Rate", min = .01, max = .2, value = .05),
  sliderInput("expRet2", "Expected Return on Risky Portfolio", min = .01, max = .6, value = .08),
  sliderInput("stDev2", "Standard Deviation on Risky Portfolio", min = .01, max = .6, value = .14)
)
renderPlot({
    y <- seq(.01, 1.5, by = .01)
   oppSet <- ifelse( y < 1, input$riskfree2 + (input$stDev2 * y) * ( (input$expRet2 - input$riskfree2) / input$stDev2), input$borrow + (input$stDev2 * y) * ( (input$expRet2 - input$borrow) / input$stDev2) )
  plot(oppSet, type = 'l', xlab = "Complete Portfolio Standard Deviaion", ylab = "Complete Portfolio Expected Return", lwd = 4, col = 3, main = "Capital Allocation Line")
  ## legend("topleft", legend = paste("Utility = ",input$utility, sep = ""), col = 3, lty = 1, lwd = 4)
})
```

## Optimal Allocation in the Risky Asset

To find the optimal place to invest on the line (the optimal value of $y$) we can plug in the equations for out complete portfolio's mean and variance into our utility function.  We then maximize the our utility function with respect to $y$.  Our utility function is:

$U = E(r_c) - \frac{1}{2}A\sigma^2_c = r_f + y(E(r_r) - r_f) - \frac{1}{2}A(y\sigma_c)^2$

Note, the function is quadratic and the $y^2$ term has a negative sign.  This means the function has a maximum and not a minimum.  So taking the derivative with respect to $y$ we have:

$\frac{dU}{dy} = E(r_r) - r_f - Ay\sigma_c^2$

setting equal to zero and solving for $y$ affords the solution:

$y^* = \frac{E(r_r) - r_f}{A\sigma_c^2}$

You can use the following to calculate the optimal $y$ given any input.

```{r calculate_y, echo=FALSE}
inputPanel(
  sliderInput("expRet3", "Expected Return", min = -.2, max = 1, value = .15),
  sliderInput("stDev3", "Standard Deviation", min = .01, max = 1, value = .2),
  sliderInput("RA3", "Degree of Risk Aversion", min = -.5, max = 6, value = 4),
  sliderInput("rf3", "Risk-free Rate", min = .01, max = .10, value = .02)
)
renderPrint({
  y = (input$expRet3 - input$rf3)/(input$RA3 * input$stDev3^2)
  out3 <- paste(round(y, 2), "%", sep = "")
  cat(out3)
})
```

#### Does this solution make sense? 

The first thing you should when you see this solution is to make sure it makes economic sense.  That is, does the proportion we invest in the risky asset $y$ change as we would expect with respect to changes in its inputs.  For example, if our degree of risk aversion ($A$) increases then we would expect $y$ to decrease.  Our solution agrees.  Check that our solution agrees with the expected behavior for $E(r_r)$, $r_f$, and $\sigma_c^2$.

<!---
### Graphical Solution for $y$

 overlay indifference curves and CAL -->

