---
title: "Portfolio Theory: Chapter 3"
author: "Matt Brigida"
date: "November 27, 2015"
output: html_document
runtime: shiny
---

In this chapter we'll construct optimal risky portfolios. We'll do so by combining risky assets according to weights that maximize the overall risky portfolio's reward-to-risk (Sharpe) ratio.  The most important aspect of this process will be that it is completely independent of the investor's degree of risk aversion.  So if two investors have the same set of risky assets, the same estimates of means and covariances for these assets, and the same restrictions on investing, they will have the same optimal risky portfolio.

This feature gave a theoretical basis for investors to delegate the construction of their risky portfolio to professional portfolio managers.  The investor then simply decides how much to invest in the professionally managed portfolio versus the risk-free asset.  Thus the construction of mean/variance optimal risky portfolios, pioneered by Harry Markowitz, had a large impact on the investment industry. This led to Markowitz receiving the Sveriges Riksbank Prize in Economic Sciences in *Memory* of Alfred Nobel in 1990.

The optimal risky portfolio is an example of *efficient* diversification.  First, we'll start with a discussion of diversification in general.

## What is Diversification?  

We are diversifying our portfolio if, given the portfolio has a set size, we split this amount across more and more assets.  Note, you are not diversifying if you add more money to your portfolio and invest it in a new asset. 


Portfolio Frontier with Two Risky Assets and Varying Correlation
=========================================================================

The following widget shows the efficient frontier for a portfolio of two risky assets.  The first risky asset (call it 'Debt') has a 9\% expected return and 11\% standard deviation.  The second portfolio (call it 'Equity') has a 12\% expected return and a 20\% standard deviation.  You are free to change the correlation coefficient between Debt and Equiry returns, and see the resulting effect on the efficient frontier.

What you should note, is that as you lower the correlation coefficient, you can receive the same expected return for less risk.  That is, investors benefit form the lower correlation.  If the correlation coefficient is -1, then you can construct a risk-free portfolio.  See below for the calculation.    


```{r, echo = FALSE}
### Efficient Frontier: 2 assets ------
#{{{
inputPanel({
    sliderInput("correl", "Correlation Coefficient", min = -1, max = 1, step = 0.01, value = 0.5, animate = TRUE)
})

renderPlot({
    ## library(ggvis)
    
    w.e <- seq(-.5,1.5, by=.01)
    w.d <- 1 - w.e
    r.e <- .12
    r.d <- .09
    E <- w.e*r.e+w.d*r.d
    s.e <- .2
    s.d <- .11
    S <- sqrt((w.e^2)*s.e^2+(w.d^2)*s.d^2+2*w.d*w.e*s.e*s.d*input$correl)
    dataEff <- data.frame(cbind(S,E,w.e))
    ## plot(S, E, type='l', xlim=c(0,.3), xlab='Portfolio Standard Deviation', ylab='Portfolio Expected Return', col = 'green')
    S.ports <- c(dataEff[dataEff$w.e == 0, ]$S, dataEff[dataEff$w.e == 1, ]$S)
    E.ports <- c(dataEff[dataEff$w.e == 0, ]$E, dataEff[dataEff$w.e == 1, ]$E)
    dataPorts <- cbind(S.ports, E.ports)
    plot(dataPorts, type='p', xlim=c(0,.3), xlab='Portfolio Standard Deviation', ylab='Portfolio Expected Return', col = 'black', ylim = c(.08, .13))
    lines(S, E, col = "green", lwd = 1.5)
    text(dataPorts, labels = c("Debt", "Equity"), cex = 1, pos = 2)

## ggvis attempt -- all looks good except for overlapping legend.    
##     effFront <- cbind(E,S, w.e)
##     effFront <- data.frame(effFront)
##     asset <- 0
##     for(i in 1:dim(effFront)[1]){
##         if(effFront$w.e[i] == 1){
##         asset[i] = "Equity" 
##     } else {
##         if(effFront$w.e[i] == 0){
##             asset[i] = "Debt" 
##         } else {
##             asset[i] = "Combined"
##         }
##     }
##     }

##     size.pt <- 0
##     for(i in 1:dim(effFront)[1]){
##         if(effFront$w.e[i] == 1){
##         size.pt[i] = 5
##     } else {
##         if(effFront$w.e[i] == 0){
##             size.pt[i] = 5 
##         } else {
##             size.pt[i] = 1
##         }
##     }
##     }
    
## effFront <- cbind(effFront, asset, size.pt)
## p <- effFront %>% ggvis(~S, ~E, size = ~size.pt) %>% layer_points(fill = ~factor(asset)) #%>% add_legend(c("fill", "size"))

##     p %>% scale_numeric("size", domain = c("Equity", "Debt", "Combined"), range = c(7, 7, 1))
})
## f <- seq(-1, 1, by = .1)
## S2 <- matrix(0, nrow=201, ncol=21)
## for (i in 1:21){Sys.sleep(.5)
##   S2[,i] <- sqrt((w.e^2)*s.e^2+(w.d^2)*s.d^2+2*w.d*w.e*s.e*s.d*f[i])
##   lines(S2[,i], E, col=i, lwd=2)}
#}}}
```

You can see if you set the coefficient to -1, then the efficient frontier has a point with no risk, and some expected return.  To calculate the expected return at that point, we first need to calculate the asset weights on the portfolio with no risk.

#### Question:  Can the risk-free rate above be different from the risk-free rate on Treasury securities?  











Testing for Portfolio Alpha
========================================================



```{r, message=FALSE,warning=FALSE, cache=TRUE}
library(quantmod)
env <- new.env()
Symbols <- c('SPY', 'QQQ', 'XLF', 'TLT','IBM','AAPL','XOM','BP','DDD')
getSymbols(Symbols = Symbols, env = env, from = '2013-01-01')
args <- eapply(env = env, FUN = function(x){ClCl(x)})[Symbols]
returns <- na.omit(do.call(what = merge, args = args))
colnames(returns) <- Symbols

## the basis for the code above is from the stackoverflow post here: (need to find and cite, note it might be quant.SE)

srets <- as.matrix(returns)
X <- cbind(1,srets[,1])
alpha <- 0
beta <- 0

for (i in 1:dim(srets)[2]) {
alpha[i] <- (solve(t(X) %*% X) %*% t(X) %*% srets[1:(dim(srets)[1]),i])[1]
beta[i] <- (solve(t(X) %*% X) %*% t(X) %*% srets[1:(dim(srets)[1]),i])[2]}

t.test(alpha)

plot(density(alpha), main = "Empirical Density Function", xlab = "alpha")

ab <- rbind(alpha,beta)

ab <- as.data.frame(ab)

names(ab) <- Symbols

attach(ab)

t.test(alpha[beta > 1],alpha[beta < 1])
```
