---
title: "Portfolio Theory: Chapter 3"
author: "Matt Brigida"
date: "November 27, 2015"
output: html_document
runtime: shiny
---

In this chapter we'll construct optimal risky portfolios. We'll do so by combining risky assets according to weights that maximize the overall risky portfolio's reward-to-risk (Sharpe) ratio.  The most important aspect of this process will be that it is completely independent of the investor's degree of risk aversion.  So if two investors have the same set of risky assets, the same estimates of means and covariances for these assets, and the same restrictions on investing, they will have the same optimal risky portfolio.

This feature gave a theoretical basis for investors to delegate the construction of their risky portfolio to professional portfolio managers.  The investor then simply decides how much to invest in the professionally managed portfolio versus the risk-free asset.  Thus the construction of mean/variance optimal risky portfolios, pioneered by Harry Markowitz, had a large impact on the investment industry. This led to Markowitz receiving the Sveriges Riksbank Prize in Economic Sciences in *Memory* of Alfred Nobel in 1990.

The optimal risky portfolio is an example of *efficient* diversification.  First, we'll start with a discussion of diversification in general.

## What is Diversification?  

Diversification is an investment strategy which reduces portfolio risk without *necessarily* reducing portfolio return.

### Diversification: A Mathematical Argument

It works because the expected return on a portfolio is the weighted-average of the expected returns of the assets in the portfolio, but the standard deviation of the portfolio is *less than* the weighted average of the individual standard deviations of the assets in the portfolio.

[Show math]

### Diversification: An Economic Argument

In economic terms, we can think of the risk on an assets as being decomposable into market and firm-specific induced risks.  Market risks (such as the risk of a sudden increse in interest rates) are common across all assets, and are not diversifiable.  Examples of firm-specific risks are a fire at a Ford factory or a drop in Facebook users.  As you add assets to your portfolio the good news form one company text to offset the bad news from another.  Adding enough assets, eventually all the firm-specific risk is offset and you hold only market risk. 

We are diversifying our portfolio if, given the portfolio has a set size, we split this amount across more and more assets.  Note, you are not diversifying if you add more money to your portfolio and invest it in a new asset. 


Portfolio Frontier with Two Risky Assets and Varying Correlation
=========================================================================

The following widget shows the efficient frontier for a portfolio of two risky assets.  The first risky asset (call it 'Debt') has a 9\% expected return and 11\% standard deviation.  The second portfolio (call it 'Equity') has a 12\% expected return and a 20\% standard deviation.  You are free to change the correlation coefficient between Debt and Equiry returns, and see the resulting effect on the efficient frontier.

What you should note, is that as you lower the correlation coefficient, you can receive the same expected return for less risk.  That is, investors benefit form the lower correlation.  If the correlation coefficient is -1, then you can construct a risk-free portfolio.  See below for the calculation.    


```{r, echo = FALSE}
### Efficient Frontier: 2 assets ------
#{{{
inputPanel({
    sliderInput("correl", "Correlation Coefficient", min = -1, max = 1, step = 0.01, value = 0.5, animate = TRUE)
})

renderPlot({
    ## library(ggvis)
    
    w.e <- seq(-.5,1.5, by=.01)
    w.d <- 1 - w.e
    r.e <- .12
    r.d <- .09
    E <- w.e*r.e+w.d*r.d
    s.e <- .2
    s.d <- .11
    S <- sqrt((w.e^2)*s.e^2+(w.d^2)*s.d^2+2*w.d*w.e*s.e*s.d*input$correl)
    dataEff <- data.frame(cbind(S,E,w.e))
    ## plot(S, E, type='l', xlim=c(0,.3), xlab='Portfolio Standard Deviation', ylab='Portfolio Expected Return', col = 'green')
    S.ports <- c(dataEff[dataEff$w.e == 0, ]$S, dataEff[dataEff$w.e == 1, ]$S)
    E.ports <- c(dataEff[dataEff$w.e == 0, ]$E, dataEff[dataEff$w.e == 1, ]$E)
    dataPorts <- cbind(S.ports, E.ports)
    plot(dataPorts, type='p', xlim=c(0,.3), xlab='Portfolio Standard Deviation', ylab='Portfolio Expected Return', col = 'black', ylim = c(.08, .13))
    lines(S, E, col = "green", lwd = 1.5)
    text(dataPorts, labels = c("Debt", "Equity"), cex = 1, pos = 2)

## ggvis attempt -- all looks good except for overlapping legend.    
##     effFront <- cbind(E,S, w.e)
##     effFront <- data.frame(effFront)
##     asset <- 0
##     for(i in 1:dim(effFront)[1]){
##         if(effFront$w.e[i] == 1){
##         asset[i] = "Equity" 
##     } else {
##         if(effFront$w.e[i] == 0){
##             asset[i] = "Debt" 
##         } else {
##             asset[i] = "Combined"
##         }
##     }
##     }

##     size.pt <- 0
##     for(i in 1:dim(effFront)[1]){
##         if(effFront$w.e[i] == 1){
##         size.pt[i] = 5
##     } else {
##         if(effFront$w.e[i] == 0){
##             size.pt[i] = 5 
##         } else {
##             size.pt[i] = 1
##         }
##     }
##     }
    
## effFront <- cbind(effFront, asset, size.pt)
## p <- effFront %>% ggvis(~S, ~E, size = ~size.pt) %>% layer_points(fill = ~factor(asset)) #%>% add_legend(c("fill", "size"))

##     p %>% scale_numeric("size", domain = c("Equity", "Debt", "Combined"), range = c(7, 7, 1))
})
## f <- seq(-1, 1, by = .1)
## S2 <- matrix(0, nrow=201, ncol=21)
## for (i in 1:21){Sys.sleep(.5)
##   S2[,i] <- sqrt((w.e^2)*s.e^2+(w.d^2)*s.d^2+2*w.d*w.e*s.e*s.d*f[i])
##   lines(S2[,i], E, col=i, lwd=2)}
#}}}
```

You can see if you set the coefficient to -1, then the efficient frontier has a point with no risk, and some expected return.  To calculate the expected return at that point, we first need to calculate the asset weights on the portfolio with no risk.

#### Question:  Can the risk-free rate above be different from the risk-free rate on Treasury securities?  



### Stock Return Correlations

You can use the following app to see a correlation matrix of the stocks' returns.  The correlation coefficients are calculted on weekly returns from 1/1/2015 to the present. You can use this to try and find stock for your optimal portfolio.  Diversification is most effective the lower the correlation between the assets. Note, the size and color of the box represents its correlation.  Can you find any stocks/assets with negative correlations (they show up as red)?

```{r correlationMatrix, echo = FALSE}
library(corrplot)
library(quantmod)
inputPanel(
    textInput("ticker1", "Stock Ticker", value = "XOM"),
    textInput("ticker2", "Stock Ticker", value = "GE"),
    textInput("ticker3", "Stock Ticker", value = "TSLA"),
    textInput("ticker4", "Stock Ticker", value = "AAPL"),
    textInput("ticker5", "Stock Ticker", value = "UNH")
)

renderPlot({
    stock1 <- getSymbols(input$ticker1, from = '2015-01-01', auto.assign = FALSE)
    stock1 <- to.weekly(stock1)
    stock1 <- Delt(Ad(stock1))[-1]
    names(stock1) <- input$ticker1
    stock2 <- getSymbols(input$ticker2, from = '2015-01-01', auto.assign = FALSE)
    stock2 <- to.weekly(stock2)
    stock2 <- Delt(Ad(stock2))[-1]
    names(stock2) <- input$ticker2
    stock3 <- getSymbols(input$ticker3, from = '2015-01-01', auto.assign = FALSE)
    stock3 <- to.weekly(stock3)
    stock3 <- Delt(Ad(stock3))[-1]
    names(stock3) <- input$ticker3
    stock4 <- getSymbols(input$ticker4, from = '2015-01-01', auto.assign = FALSE)
    stock4 <- to.weekly(stock4)
    stock4 <- Delt(Ad(stock4))[-1]
    names(stock4) <- input$ticker4
    stock5 <- getSymbols(input$ticker5, from = '2015-01-01', auto.assign = FALSE)
    stock5 <- to.weekly(stock5)
    stock5 <- Delt(Ad(stock5))[-1]
    names(stock5) <- input$ticker5
M <- cor(data.frame(cbind(stock1, stock2, stock3, stock4, stock5)))
corrplot(M, method = "square")
})
```

### Downloading Stock Returns

You can use the following app to download stock returns.  Input the stock tickers in the text boxes and choose the your time interval. Once you see the data in the table you can click the 'Download' button to download a csv file with the returns.  If you use this to download multiple sets of stock, note that the downloaded file is named 'stockReturns.csv' and you may want to change the name of preiously downloaded files to avoid them being overwritten.


```{r stockRetDownload, echo = FALSE}
library(quantmod)
library(DT)
inputPanel(
    textInput("ticker21", "Stock Ticker", value = "XOM"),
    textInput("ticker22", "Stock Ticker", value = "GE"),
    textInput("ticker23", "Stock Ticker", value = "TSLA"),
    textInput("ticker24", "Stock Ticker", value = "AAPL"),
    textInput("ticker25", "Stock Ticker", value = "UNH"),
    dateRangeInput("dates", "Input a Date Range", start = "2015-01-01", end = Sys.Date()),
    downloadButton('downloadData', 'Download')
)

DT::renderDataTable({
    stock1 <- getSymbols(input$ticker21, from = input$dates[1], to = input$dates[2], auto.assign = FALSE)
    stock1 <- to.weekly(stock1)
    stock1 <- Delt(Ad(stock1))[-1]
    names(stock1) <- input$ticker21
    stock2 <- getSymbols(input$ticker2, from = input$dates[1], to = input$dates[2], auto.assign = FALSE)
    stock2 <- to.weekly(stock2)
    stock2 <- Delt(Ad(stock2))[-1]
    names(stock2) <- input$ticker22
    stock3 <- getSymbols(input$ticker3, from = input$dates[1], to = input$dates[2], auto.assign = FALSE)
    stock3 <- to.weekly(stock3)
    stock3 <- Delt(Ad(stock3))[-1]
    names(stock3) <- input$ticker23
    stock4 <- getSymbols(input$ticker4, from = input$dates[1], to = input$dates[2], auto.assign = FALSE)
    stock4 <- to.weekly(stock4)
    stock4 <- Delt(Ad(stock4))[-1]
    names(stock4) <- input$ticker24
    stock5 <- getSymbols(input$ticker5, from = input$dates[1], to = input$dates[2], auto.assign = FALSE)
    stock5 <- to.weekly(stock5)
    stock5 <- Delt(Ad(stock5))[-1]
    names(stock5) <- input$ticker25
M2 <<- data.frame(cbind(stock1, stock2, stock3, stock4, stock5))
    M2
})

output$downloadData <- downloadHandler(
    filename = function() { paste("stockReturns", '.csv', sep='') },
    content = function(file) {
      write.csv(M2, file)
    }
  )
```


Testing for Portfolio Alpha
========================================================



```{r, message=FALSE,warning=FALSE, cache=TRUE}
library(quantmod)
env <- new.env()
Symbols <- c('SPY', 'QQQ', 'XLF', 'TLT','IBM','AAPL','XOM','BP','DDD')
getSymbols(Symbols = Symbols, env = env, from = '2013-01-01')
args <- eapply(env = env, FUN = function(x){ClCl(x)})[Symbols]
returns <- na.omit(do.call(what = merge, args = args))
colnames(returns) <- Symbols

## the basis for the code above is from the stackoverflow post here: (need to find and cite, note it might be quant.SE)

srets <- as.matrix(returns)
X <- cbind(1,srets[,1])
alpha <- 0
beta <- 0

for (i in 1:dim(srets)[2]) {
alpha[i] <- (solve(t(X) %*% X) %*% t(X) %*% srets[1:(dim(srets)[1]),i])[1]
beta[i] <- (solve(t(X) %*% X) %*% t(X) %*% srets[1:(dim(srets)[1]),i])[2]}

t.test(alpha)

plot(density(alpha), main = "Empirical Density Function", xlab = "alpha")

ab <- rbind(alpha,beta)

ab <- as.data.frame(ab)

names(ab) <- Symbols

attach(ab)

t.test(alpha[beta > 1],alpha[beta < 1])
```

###
