---
title: "Portfolio Theory: Chapter 7 (The Pairs Trade)"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Authors: [GitHub Contributors](https://github.com/Matt-Brigida/portfolio-theory/graphs/contributors)

The pairs trade is a classic example of a statistical arbitrage (stat arb) strategy---its history dates back to the 1980s.  

This stragegy can be thought of as the oppposite of Markowitz mean/variance optimization.  In Markowitz we diversified efficiently, which reduces firsm specific risk but leaves market risk in our portfolio.  In the pairs trade, we'll remove market risk, and hold only firm-specific risk. Thus, the pairs trade is known as a 'market-neutral' strategy. 


## Implementation

There is no single way to implement the pairs trade. This is an active area of research, and provides you with a venue to create your own strategy (and become wealthy if it works).  However, to learn we'll implement the strategy outlined in Gatev et al ().  


To implement the pairs trade we'll first define what is normal behavior for a particular pair.  We do this over the 'estimation period'.  

# Estimation Period

```{r, messsage=FALSE, warning=FALSE, echo = FALSE}
## in the process of translating this code into a Shiny app
## just run for each pair?

inputPanel(
  dateRangeInput("estim_period", "Estimation Period", start = "2015-09-01", end = "2015-12-31"),
  ## selectizeInput(tickers, "Enter Tickers", )
  textInput("ticker1", "First Stock Ticker", value = "XOM"),
  textInput("ticker2", "Second Stock Ticker", value = "COP")
)

renderPrint({
    
library(quantmod)
## Estimation period
from <- input$estim_period[1]
to <- input$estim_period[2]

## The code below to calculate an xts object of returns is from an SO or Quant.SE post.  I need to find it and cite.
env <- new.env()
Symbols <- c(input$ticker1, input$ticker2)
getSymbols(Symbols = Symbols, env = env, from = from, to = to)
args <- eapply(env = env, FUN = function(x){ClCl(x)})[Symbols]
returns <- na.omit(do.call(what = merge, args = args))
colnames(returns) <- Symbols
## convert reutrns to  normalized prices
norm.prices <- cumsum(returns)

norm.prices.df <- as.data.frame(norm.prices)
## function to find the average squared deviation of the normalized prices for two stocks
## stats <- function(TIC1, TIC2){
    cat("The average squared deviation is", mean(( norm.prices.df[,1] - norm.prices.df[,2])^2 ),"\n")
    cat("The standard deviation of the squared deviations is:", sd(( norm.prices.df[,1] - norm.prices.df[,2])^2 ),"\n")
## }

## run this function by typing g() into the interpreter and it will create
## a .csv file in your working directory with all the pairs and their
## avg squared dev and stdev of squared deviation sorted from lowest
#  avg sq dev to highest

## g <- function(){
## list <- matrix(0,nrow=dim(norm.prices.df)[2], ncol=dim(norm.prices.df)[2])
## asds <- matrix(0,nrow=dim(norm.prices.df)[2], ncol=dim(norm.prices.df)[2])
## asstdevs <- matrix(0,nrow=dim(norm.prices.df)[2], ncol=dim(norm.prices.df)[2])
## for(i in 1:dim(norm.prices.df)[2]){
##     for(j in 1:dim(norm.prices.df)[2]){
##         list[i,j] <- as.character(paste(names(norm.prices.df)[i],names(norm.prices.df)[j]),sep="")
##         asds[i,j] <- mean((norm.prices.df[,i]-norm.prices.df[,j])^2)
##         asstdevs[i,j] <- sd((norm.prices.df[,i]-norm.prices.df[,j])^2)
##           }
##     }

})

```

```{r echo = FALSE}
## renderPlot({
## ll <- dim(norm.prices.df)[2]
## tmp1 <- rep(99,times=ll^2)
## tmp2 <- rep(99,times=ll^2)
## tmp3 <- rep(99,times=ll^2)
##  for(z in 0:(ll-1)){
##  tmp1[(z*ll+1):(z*ll+ll)] <- list[(z+1),]
##  tmp2[(z*ll+1):(z*ll+ll)] <- asds[(z+1),]
##  tmp3[(z*ll+1):(z*ll+ll)] <- asstdevs[(z+1),]
## }
## est.stats <- cbind(tmp1, tmp2, tmp3)
## est.stats <- as.data.frame(est.stats)
## names(est.stats) <- c("Pair","Avg. Squared. Dev","StdDev Squared Dev")
## est.stats <- est.stats[est.stats$'Avg. Squared. Dev'!=0,]
## est.stats <- est.stats[with(est.stats, order(est.stats[,2])),]
## est.stats <- est.stats[seq(2, nrow(est.stats),by=2),]
## write.csv(est.stats, file='stats_forall_pairs.csv')
## }
## })
```

<!--- end -->
