---
title: "Portfolio Theory: Chapter 1"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#### Authors: [GitHub Contributors](https://github.com/Matt-Brigida/portfolio-theory/graphs/contributors)

<!---
Note, much of the R code in this text was written long ago, when I wrote even worse R code than I do now.  The code works, but it is inefficient and ugly.  I decided not to rewrite it however, feeling it more important to get the text out and into students' hands.  If you can improve the code, please feel free.  Pull requests are always welcome.
-->
*The text and code for this book is [hosted on GitHub](https://github.com/Matt-Brigida/portfolio-theory).  If you have improvements, please feel free to send along pull requests.*

# Introduction

Despite 'theory' in the title, this course will teach you how to construct portfolios in the real world.  We'll generally introduce a topic, learn the theoretical underpinnings, and then try to apply the topic to real data.  Some of the most interesting topics in this course will be on the implementation of the theories we have learned.  

Through this course the student will gain a working knowledge of what we mean by the term *risk* in finance.  There will be differing measures of risk depending on the assumptions you make about the world.  Then, for a given set of assumptions, risk will directly determine portfolio allocations.  We'll also cover an extension of this, where the resulting portfolio allocations will imply an equilibrium expected return for each asset (the CAPM).  An apt alternative title for this course would be 'Risk and its Implications'. 

### Arithmetic versus Geometric Returns

Given a time series of returns, we can calculate arithmetic and geometric average returns.  Each type of average return (arithmetic vs geometric) is trying to answer a different question.

#### Arithmetic average:

You may recall, if we are given a discrete probability density function which describes returns, we can calculate the expected, or average, return as $E(r) = \sum_{s=1}^N{p(s)r(s)}$ where $s$ denotes the state of the world, and $p(s)$ and $r(s)$ is the probability of that state, and the return in that state, respectively.

When we switch to a time series of historical returns, we assign each return observation as equally likely.  So if there are $T$ returns, then each has probability $\frac{1}{T}$.  Therefore, the arithmetic average return is:

$\bar{r} = \frac{1}{T}\sum_{t = 1}^T{r_t}$

The arithmetic average return may be interpreted as the return we expect if we were to choose one year's return at random.  Also, if we assume the future is like the past, then we can say, given no other information, it is the return we expect in any given future year.   

#### Geometric average:

The geometric average (denote by $GA$) return is:

$GA = \left(\prod_{t=1}^T(1 + r_t)\right)^{\frac{1}{T}} - 1$

In other words, add 1 to each return, multiply all the returns together, and take the $T$th root of the product of the returns.  Then subtract 1, and you have the geometric average.  The geometric average can be interpreted as the average return per period you expect to earn over a set period.  

As a note, the geometric average is always less than or equal to the arithmetic average. The only case where the averages are equal is of all the returns being averaged are equal.  Can you prove this?  



### Real vs Nominal Rates

In what follows we'll be mainly concerned with nominal rates of return.  The *nominal* rate is the change in the amount of currency (US dollars) you have. Conversely, the *real* rate of return is the change in your purchasing power---the amount of stuff you can buy.  Since currency is only (1) a medium of exchange (for stuff) and a (2) store of value, we really only care about our real rate of return.

However, we can never really know what our real return was, because to calculate it we need the nominal rate and inflation.  We know the nominal rate with certainty, however we can only estimate what inflation was (or will be).  Therefore, there is some uncertainty as to the value of the real rate.  Below is a brief table highlighting some salient differences between the real and nominal rates. Note, $r$ is the real rate, $R$ the nominal, and $i$ denotes inflation. 

|        | Real Rate | Nominal Rate |
|--------|-----------|--------------|
| Means  | Change in what you can buy| Change in dollars |
| Certainty | We only estimate the real rate | We can observe the nominal rate |
| Solved for: | $r = \frac{R -i}{1 + i}$ | $R = r + E(i)$ |
| Taxes | Taxes are not paid on the real rate | Taxes are paid on the nominal rate |
| As an investor | You are concerned with a high (risk-adjusted) after-tax real rate of return | Only relevant as it affects real rate |
|    |     |      |


### Risk Premia

An assets risk premium is the reward you expect to earn for bearing risk.  We can express the risk premium as $E(r) - r_f$ where $E(r)$ is the expected return on a risky asset, and $r_f$ is the risk free rate.  Understanding risk premia will be a central part of this course.

### The Standard Deviation and Variance

A random variable's (here returns) variance is its expected squared deviation from its mean.  Given a discrete distribution which describes returns we then have:

$\sigma^2 = \sum_{s=1}^Sp_s(r_t - E(r))^2$

where $\sigma^2$ denotes the variance, and $r_s$ and $p_s$ denote the return in state $s$, and the probability of state $s$ respectively.  If we were instead to assume a continuous probability density function (such as the normal distribution), then the variance would be:

$\sigma^2 = \int {(r - E(r))^2f(r)dr}$

This is more simple than it looks.  $f(r)dr$ is the same rectangle as $p_s$ in the discrete version.  The only different is $f(r)dr$ is in the limit where the base of the rectangle gets smaller and smaller.  Note, you won't need to know the distinctions between the discrete and continuous cases for what follows---I just thought you should see both equations in case they come up in some context later.

In the cast of historical data, with $T$ observations, the variance is:

$\hat{\sigma}^2 = \frac{1}{T-1}\sum_{t=1}^T(r_t - \bar{r})^2$

where each period's deviation from the mean is given equal weight ($\frac{1}{T-1}$).  We divide by $\frac{1}{T-1}$ instead of $\frac{1}{T}$ to adjust for the one degree of freedom we lose in calculating the mean.

Since the variance is the average *squared* deviations, the value is not in the units of the variable for which we are calculating the variance.  In other words, when we calculate the variance of a stock's percent returns, the result is not in percent returns.

To return to our original units (percent returns) we take the square root of the variance.  The positive square root of the variance is the *standard deviation*.  This is in the original units---percent returns.  

$\hat{\sigma} = \sqrt{\frac{1}{T-1}\sum_{t=1}^T(r_s - \bar{r})^2}$

### The Sharpe Ratio

The Sharpe Ratio is the ratio of an investment's reward to is volatility. In other words, it is what you receive in excess of the risk-free rate scaled by the risk you have to take to get that return.

$Sharpe\ Ratio=\frac{Risk\ Premium}{Volatility}$

where the volatility is defined as the standard deviation of excess returns.

<!---insert app which will calculate historical average return, variance, standard deviation, skewness, kurtosis, and sharpe ratios-->

## The Normal Distribution

Very often in finance we assume returns are normally distributed.  This assumption means, given a mean and variance, the distribution is completely determined.  That is, is an asset's returns are assumed normal, then all we need to fully describe the asset is its mean and variance.  This effectively assumes the values of [higher moments](https://en.wikipedia.org/wiki/Moment_\(mathematics\)#Higher_moments) (such as skewness and kurtosis).  

If we assume a normal distribution, then we can use the standard deviation as a measure of risk.  When we refer to risk, we mean the probability of *loss*---the left tail of the return distribution.  These are the returns below zero (or some benchmark).  However, since the normal distribution is symmetric, the left and right tails are the same.  So it doesn't matter that the standard deviation is averaging both the deviations above and below the mean.  If the distribution is not symmetric, then the deviations below the mean may be quite different from the deviations above the mean, and the standard deviation makes little sense as a measure of risk.  

Also, the normal distribution has the useful feature that the [sum of normally distributed random variables is also normally distributed](https://en.wikipedia.org/wiki/Sum_of_normally_distributed_random_variables).  Note, also a [constant times a normal distribution is normally distributed](http://math.stackexchange.com/questions/275648/multiplication-of-a-random-variable-with-constant).  This means when we calculate a portfolio's return it will also be normally distributed if we assume each asset in the portfolio is normally distributed.  Note, the portfolio return is the sum of the return on each asset times its weight in the portfolio, but more on that later.

## Deviations from Normality

Assuming asset returns is convenient, however is it plausible?  Moreover, in what ways can assets returns differ from the normal distribution?  Below are interactive applications which will allow you to investigate these questions yourself.  Before you do, however, we'll take a brief look at ways in which returns may deviation from normality.

### Skewness

Skewness is a measure of asymmetry in the distribution.  It is $\frac{E[(r - E(r))^3]}{E([(r - E(r))^2])^{3/2}}$.  It is not important to remember this necessarily, however do note that the deviation af the asset return from its mean is raised to the power of 3, which retains the sign of the deviation.  In other words gains and losses are treated differently.

Positively skewed assets have long tails in the direction of positive returns, and the opposite for negatively skewed assets. As mentioned earlier, symmetric measures of risk such as the standard deviation are less relevant if the distribution is very skewed on either direction.

There is ample evidence that asset return distributions are skewed, particularly for individual assets as opposed to portfolios.  Some of this evidence points to [risk-takers actively seeking out and buying positively skewed stocks](https://scholar.google.com/scholar?cluster=3640748709756050519&hl=en&as_sdt=0,39). Conversely, negatively skewed assets would want to be avoided (unless of course if you intend to short).

### Kurtosis

Kurtosis is a measure of the fatness of the tails of a distribution---how likely there are to be outlier events. It is $\frac{E[(r - E(r))^2]}{E([(r - E(r))^2])^2}$.  Again, remembering this is not important, but notice the assets deviation from its expected value is squared.  So we lost the sign of the deviation.  Gains and losses are treated the same.

The normal distribution has a kurtosis of three, so often people will quote the *excess* kurtosis, which is the calculated kurtosis less three.  It is worthwhile to remember that when using a built-in function to calculate kurtosis, some software will return the kurtosis, and some will return the excess kurtosis.  Be sure to read the appropriate documentation before using the function.

### How good of an approximation is the normal distribution?

The idea of the application below is that in what follows we'll assume that stock behaves according to the normal distribution (red dotted line).  The empirical distribution is how the stock actually behaved over the interval in which we calculated the stock's parameters (mean and standard deviation).

Try different stocks and start dates.  You'll generally find small stocks over short time intervals have very non-normal empirical distributions.  However, often large stocks over long periods will have non-normal distributions as well.  The question is, given this evidence, why do we assume the normal distribution, and is it incorrect to do so?  

```{r comparetoNormal, echo=FALSE}
## compare normal and empirical distributions 
#{{{
inputPanel(
  textInput("ticker1", "Stock Ticker", value = "GE"),
  dateInput("startDate1", "Start Date", value = "2015-01-01")
)
```
```{r, echo = FALSE}
renderPlot({
library(tseries)
library(quantmod)
  
    x <- get.hist.quote(input$ticker1, start = input$startDate1, quote = "Close")
    x <- xts(x)
    r <<- Delt(x, type = 'log')[-1]

    plot(density(r), main = "Normal versus Empirical Distribution", lty = 1, 
        col = 1, lwd = 2, xlab = "Log-Returns in %/100") #, sub = "Matthew Brigida; Clarion UofP")

    lines(density(rnorm(5e+05, mean = mean(r), sd = sd(r))), lty = 4, col = 2, 
        lwd = 2)

    legend("topright", c("Stock Return Density", "Normal Density"), col = c(1, 2), 
           lty = c(1, 4), lwd = c(2, 2))
})
```
```{r, echo = FALSE}
library(tseries)

renderPrint({
        null <- input$ticker1
        null <- input$startDate1
    cat("The sample skewness is", skewness(r), "\n")
    cat("For a t-statistic of", skewness(r)/(sqrt(6/length(r))), "\n")
    p1 <- 2 * (1 - pt(abs(skewness(r)/(sqrt(6/length(r)))), length(r) - 1))
    cat("And a p-value of", p1, "\n")
    cat("So we", ifelse(p1 < 0.05, "reject the null, and find the distribution is skewed.", 
        "do not reject the null, the distribution is symmetric."), "\n")
    cat("\n")
    
    cat("The sample excess kurtosis is", kurtosis(r)[1], "\n")
    cat("For a t-statistic of", kurtosis(r)/(sqrt(24/length(r))), "\n")
    p2 <- 2 * (1 - pt(abs(kurtosis(r)/(sqrt(24/length(r)))), length(r) - 1))
    cat("And a p-value of", p2, "\n")
    cat("So we", ifelse(p2 < 0.05, "reject the null, and find the distribution has fat tails.", 
        "do not reject the null, the distribution does not have fat tails."), 
        "\n")
                                        cat("\n")
        
})    
#}}}
```


<!--- commented until done
## Risk and Return Over our History

Most texts in portfolio theory contain a summary of the rate of return on different asset classes over time.  This is to show you different assets have different characteristics---on the whole assets with greater risk have higher long-term returns.  We'll be no different. However, take care to note the point of this exercise is simply to get an intuitive feel of the idea that taking more risk grants a higher return.

Further, it can't be mentioned enough that ultimately it will be the covariance (correlation) between asset which will greatly affect the weight of the asset in our portfolio, and the expected return on that asset.  So take care to not how one asset behaves relative to another.  

Note, throughout this text we'll generally quote returns on an annual basis.  This is important, because to compare returns between any two investments, we must look at the returns over the same length of time.  Beware anyone who quotes you their return over (particularly) longer periods than a year, "I doubled my money in 10 years".  This is somewhat deceiving because a 100\% return in 10 years is only 7.18\% per year.

```{r riskReturnHistory, echo=FALSE, message = FALSE, warning = FALSE, cache = FALSE, eval = FALSE}
library(shiny)
library(dygraphs)
library(quantmod)
renderPlot({

    tenYr <- getSymbols('GS10', src = 'FRED')
    threeMonth <- getSymbols('TB3MS', src = 'FRED')
    moodysAAA <- getSymbols('AAA', src = 'FRED')
    moodysBAA <- getSymbols('BAA', src = 'FRED')
    djiaTo1968 <- getSymbols('M1109BUSM293NNBR', src = 'FRED')
    


})
```

### A Brief Note on Ergodicity 

You may point out that what we have just done is not really compelling evidence. We have looked at an historical sample of size 1, and implied that this must hold.  In fact, to make any sort of statement about the empirical relationship between risk and return we would need to sample many times from all possible histories, and then average over these histories.  Of course, we can only ever observe one history---an important distinction from empirical results in disciplines such as physics and chemistry.

However, if the time series are ergodic, then averaging over time gives the same result as averaging over possible histories. The very interested student can see Hamilton's *Time Series Analysis*.  

Later, however we'll also see theories which imply a similar risk/return relationship. This will lend credence to these earlier empirical observations.  

-->

# Risk Measures

A very active area of research in finance concerns ways to evaluate and express risk.  The student should know, and be encouraged, by the fact that there is still a great deal of contributions that can be made in here.  We'll briefly discuss the main measure, VaR.  There are some other measures, such as Conditional Tail Expectation which the interested reader may want to consider.  

## Value at Risk (VaR)

VaR is stated:  $VaR_\alpha = inf\{x \in \mathbb{R} : P(L > x) \leq 1 - \alpha\}$, where $\alpha$ is the confidence level (such as 95\% or 99\%), and $L$ is some loss.  In words this reads the smallest real number $x$ such that the probability that the loss is greater than $x$ is less than 1\% (or 5\% depending on the confidence level).

So say we are using a 99\% confidence level, and we are using daily data, and we find $x = \$500,000$.  Then we can state this like, `There is a 1\% or less chance on any given day that we will lose more than \$500,000'.   

This measure of risk makes sense, and accordingly has been written into regulations concerning levels of risk banks may take. However, the difficulty is in its implementation.  We need some way to describe $P(L > x)$.  We can assume a normal distribution, or any other we would like. In addition to this assumption, we have to come up with the distribution's parameters (such as mean and variance).  Do we use historical data, or a other forecasting method?  Or we can use historical quantiles or empirical density functions.  In this fashion we *only* use history, without any assumption of a distribution.

None of these methods are satisfactory.  Using historical data assumes the future is like the past.  This is why it is often said VaR works until you need it to work---in a market crash the future is distinctly unlike the past.  We could try and forecast the future, but no one has been able to do this well and consistently. Assuming a normal distribution, whether we use parameters based on historical data or forecasts, is really the least of the problems with implementing VaR (or other risk measures).  

One of the largest benefits of using something like VaR is simply that it forces you to think about your exposures to different risks.  In practice it gives a structure to your risk management thought process.  Also, note, a VaR calculation is not simply done once, but rather is an iterative procedure.  You calculate your VaR, improve assumptions, measures, and adjust for changing market conditions, and recalculate. 

### Interactive VaR Application

These functions output VaR estimates using both the historical empirical density function and a normal density (with a mean and standard deviation based on the historical data).  They both assume you have \$1 million invested in the asset. 

Enter a ticker and see the resulting 1\% and 5\% VaR be using both a normal distribution and [empirical density function](https://en.wikipedia.org/wiki/Empirical_distribution_function).  Note VaR makes more sense with respect to a portfolio, so try some [ETFs](https://en.wikipedia.org/wiki/Exchange-traded_fund) like SPY and EWZ.  

#### Using a 5\% VaR

```{r var5, echo=FALSE}
inputPanel(
    textInput("tickerVar5", "Input a Stock Ticker", "XOM"),
    dateInput("startDateVar5", "Start date:", value = "2015-01-01")
)

renderPrint({
    library(fBasics)
    library(tseries)
    x <- get.hist.quote(input$tickerVar5, start = input$startDateVar5, quote="Close")
x <- ts(x)
r <- log(x[2:(length(x))]/x[1:(length(x)-1)])
emp.var <- quantile(r, probs=0.05)*1000000
cat("The empirical VaR is", emp.var, "\n")
cat("The VaR assuming a normal distribution is", quantile(rnorm(500000, mean=mean(r), sd=sd(r)), probs=0.05)*1000000, "\n")
})
```

#### Using a 1\% VaR

```{r var1, echo = FALSE}
inputPanel(
    textInput("tickerVar1", "Input a Stock Ticker", "XOM"),
    dateInput("startDateVar1", "Start date:", value = "2015-01-01")
)

renderPrint({
    library(fBasics)
    library(tseries)
    x <- get.hist.quote(input$tickerVar1, start = input$startDateVar1, quote="Close")
x <- ts(x)
r <- log(x[2:(length(x))]/x[1:(length(x)-1)])
emp.var <- quantile(r, probs=0.01)*1000000
cat("The empirical VaR is", emp.var, "\n")
cat("The VaR assuming a normal distribution is", quantile(rnorm(500000, mean=mean(r), sd=sd(r)), probs=0.01)*1000000, "\n")
})
```

<!-- end -->

